---
title: "Class19: Pertussis and the CMI-PB project "
author: "Aisha"
format: pdf
---

Pertussis is a severe lung disease that is also known as whopping cough.

We will begin by investigating the number of Pertussis cases per year in the US. 

The data is linked here: (https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html) 

```{r}
echo = FALSE

cdc <- data.frame(
                                                                                                                                                                                                            Year = c(1922L,1923L,1924L,
                                                                                                                                                                                                                     1925L,1926L,1927L,
                                                                                                                                                                                                                     1928L,1929L,1930L,
                                                                                                                                                                                                                     1931L,1932L,1933L,
                                                                                                                                                                                                                     1934L,1935L,1936L,
                                                                                                                                                                                                                     1937L,1938L,1939L,
                                                                                                                                                                                                                     1940L,1941L,
                                                                                                                                                                                                                     1942L,1943L,1944L,
                                                                                                                                                                                                                     1945L,1946L,1947L,
                                                                                                                                                                                                                     1948L,1949L,1950L,
                                                                                                                                                                                                                     1951L,1952L,1953L,
                                                                                                                                                                                                                     1954L,1955L,1956L,
                                                                                                                                                                                                                     1957L,1958L,1959L,
                                                                                                                                                                                                                     1960L,1961L,
                                                                                                                                                                                                                     1962L,1963L,1964L,
                                                                                                                                                                                                                     1965L,1966L,1967L,
                                                                                                                                                                                                                     1968L,1969L,1970L,
                                                                                                                                                                                                                     1971L,1972L,1973L,
                                                                                                                                                                                                                     1974L,1975L,1976L,
                                                                                                                                                                                                                     1977L,1978L,1979L,
                                                                                                                                                                                                                     1980L,1981L,
                                                                                                                                                                                                                     1982L,1983L,1984L,
                                                                                                                                                                                                                     1985L,1986L,1987L,
                                                                                                                                                                                                                     1988L,1989L,1990L,
                                                                                                                                                                                                                     1991L,1992L,1993L,
                                                                                                                                                                                                                     1994L,1995L,1996L,
                                                                                                                                                                                                                     1997L,1998L,1999L,
                                                                                                                                                                                                                     2000L,2001L,
                                                                                                                                                                                                                     2002L,2003L,2004L,
                                                                                                                                                                                                                     2005L,2006L,2007L,
                                                                                                                                                                                                                     2008L,2009L,2010L,
                                                                                                                                                                                                                     2011L,2012L,2013L,
                                                                                                                                                                                                                     2014L,2015L,2016L,
                                                                                                                                                                                                                     2017L,2018L,2019L,
                                                                                                                                                                                                                     2020L,2021L),
                                                                                                                                                                                    Cases = c(107473,164191,165418,
                                                                                                                                                                                                                     152003,202210,
                                                                                                                                                                                                                     181411,161799,197371,
                                                                                                                                                                                                                     166914,172559,
                                                                                                                                                                                                                     215343,179135,265269,
                                                                                                                                                                                                                     180518,147237,
                                                                                                                                                                                                                     214652,227319,103188,
                                                                                                                                                                                                                     183866,222202,
                                                                                                                                                                                                                     191383,191890,109873,
                                                                                                                                                                                                                     133792,109860,
                                                                                                                                                                                                                     156517,74715,69479,
                                                                                                                                                                                                                     120718,68687,45030,
                                                                                                                                                                                                                     37129,60886,62786,
                                                                                                                                                                                                                     31732,28295,32148,
                                                                                                                                                                                                                     40005,14809,
                                                                                                                                                                                                                     11468,17749,17135,
                                                                                                                                                                                                                     13005,6799,7717,9718,
                                                                                                                                                                                                                     4810,3285,4249,
                                                                                                                                                                                                                     3036,3287,1759,
                                                                                                                                                                                                                     2402,1738,1010,2177,
                                                                                                                                                                                                                     2063,1623,1730,
                                                                                                                                                                                                                     1248,1895,2463,
                                                                                                                                                                                                                     2276,3589,4195,2823,
                                                                                                                                                                                                                     3450,4157,4570,
                                                                                                                                                                                                                     2719,4083,6586,
                                                                                                                                                                                                                     4617,5137,7796,6564,
                                                                                                                                                                                                                     7405,7298,7867,
                                                                                                                                                                                                                     7580,9771,11647,
                                                                                                                                                                                                                     25827,25616,15632,
                                                                                                                                                                                                                     10454,13278,16858,
                                                                                                                                                                                                                     27550,18719,48277,
                                                                                                                                                                                                                     28639,32971,
                                                                                                                                                                                                                     20762,17972,18975,
                                                                                                                                                                                                                     15609,18617,6124,
                                                                                                                                                                                                                     2116)
                                                                                                                                                                                  )
```


```{r}
head(cdc)
```

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data
frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
library(ggplot2)

ggplot(cdc) + aes(Year, Cases) + geom_point() + geom_line() + labs(title = "Pertussis Cases by Year (1992-2019)")
```

> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of 
the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}

ggplot(cdc) + aes(Year, Cases) + geom_point() + geom_line() + 
  labs(title = "Pertussis Cases by Year (1992-2019)") + 
  geom_vline(xintercept = 1946, linetype = "dashed", col = "blue") + 
  geom_vline(xintercept = 1996, linetype = "dashed", col = "red") + 
  geom_vline(xintercept = 2019, linetype = "dashed", col = "green") 
```

We can see a drop of cases after the introduction of the 46' vaccine and a small increase with the 96'
vaccine. 


> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation 
for the observed trend?

After the introduction of the aP vaccine, there is a small lag and then we can see how the Pertussis 
cases have increasingly rose afterwards. The aP vaccine doesn't seem to be as effective/long-term such 
as the wP vaccine. 


# Exploring CMI-PB data

Why is this vaccine-preventable disease on the upswing? To answer this question we need to investigate
the mechanisms underlying waning protection against pertussis. This requires evaluation of 
pertussis-specific immune responses over time in wP and aP vaccinated individuals.

This is the goals of the CMI-PB project: https://www.cmi-pb.org/

The CMI-PB project makes its data available via "API-endpoint" that return the JSON format. 

We will use the **jsonlite** package to access this data. The main funciton in this package is 
called `read_json()`. 


```{r}
library(jsonlite)

# Subject table

subject <- read_json("http://cmi-pb.org/api/subject", simplifyVector = TRUE)
specimen <- read_json("http://cmi-pb.org/api/specimen", simplifyVector = TRUE)

titer <- read_json("http://cmi-pb.org/api/v4/plasma_ab_titer", simplifyVector = TRUE)

```

Lets have a look at these new objects: 

```{r}
head(subject)
```


```{r}
head(titer)
```

> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```

```{r}
library(tidyverse)
```

```{r}
today()
```

```{r}
today() - mdy("11-28-2001")
```

```{r}
time_length( today() - ymd("2002-08-18"), "years")
```


> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age 
of aP individuals; and (iii) are they significantly different?

```{r}
subject$age <- today() - ymd(subject$year_of_birth)
```

```{r}
library(dplyr)

 ap <- subject %>% filter(infancy_vac == "aP")

round( summary( time_length( ap$age, "years" ) ) )
```

```{r}
# wP

wp <- subject %>% filter(infancy_vac == "wP") 
round( summary( time_length( wp$age, "years" ) ) )
```

The difference between both of their average ages is about 10 years. 

> Q8. Determine the age of all individuals at time of boost?

```{r}
subject$age <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)

subject$age_year <- time_length(subject$age, "year")

head(subject)
```

> Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups 
are significantly different?

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```

These two groups are very significantly differently. 

# Merge or join tables


> Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all
specimen records along with their associated subject details:

```{r}
meta <- inner_join(specimen, subject)

dim(meta)
head(meta)
```

Antibody measurements in the blood. 

```{r}
head(titer)
```

> Q10. Now using the same procedure join meta with titer data so we can further analyze this data in 
terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(titer, meta)

dim(abdata)
head(abdata)
```
> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```


> Q12. What are the different $dataset values in abdata and what do you notice about the number of 
rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```

The number of rows for the 2022_data is far less than the other datasets. The number of rows seems
to decrease as the years go by. 


# Examine IgG Ab titer levels 


Let's focus on one of these IgG. 

```{r}
igg <- abdata %>% filter(isotype == "IgG")

head(igg)
```

Boxplot of MFI_normalised vs antigen. 

```{r}
ggplot(igg) + aes(MFI_normalised, antigen) + geom_boxplot()
```

```{r}
ggplot(igg) + 
  aes(MFI_normalised, antigen, col = infancy_vac) + 
  geom_boxplot()

```

> Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) + 
  aes(MFI_normalised, antigen) + 
  geom_boxplot() + xlim(0,75) + facet_wrap(vars(visit), nrow=2)
```

```{r}
head(igg)
```


Focus in on IgG to the Pertussis Toxin (PT) antigen. 


```{r}
igg.pt <- igg %>% filter(antigen == "PT", dataset == "2021_dataset")
```

```{r}
ggplot(igg.pt) + 
  aes(planned_day_relative_to_boost, MFI_normalised, 
      col = infancy_vac, group = subject_id) + 
  geom_point() + geom_line() + geom_vline(xintercept = 0, linetype = "dashed", col = "black") + 
  geom_line() + geom_vline(xintercept = 14, linetype = "dashed", col = "black") 
```



# Obtaining CMI-PB RNASeq data

For RNA-Seq data the API query mechanism quickly hits the web browser interface limit for file size. 
We will present alternative download mechanisms for larger CMI-PB datasets in the next section. However, 
we can still do “targeted” RNA-Seq querys via the web accessible API.

```{r}
url <- "https://www.cmi-pb.org/api/v2/rnaseq?versioned_ensembl_gene_id=eq.ENSG00000211896.7"

rna <- read_json(url, simplifyVector = TRUE) 
```


To facilitate further analysis we need to “join” the rna expression data with our metadata meta, which 
is itself a join of sample and specimen data.

```{r}
#meta <- inner_join(specimen, subject)
ssrna <- inner_join(rna, meta)
```

> Q19. Make a plot of the time course of gene expression for IGHG1 gene (i.e. a plot of visit vs. tpm).

```{r}
ggplot(ssrna) +
  aes(x = visit, y = tpm, group=subject_id) +
  geom_point() +
  geom_line(alpha=0.2)
```

> Q20.: What do you notice about the expression of this gene (i.e. when is it at it’s maximum level)?

The expression of this gene when it is at its' maximum level has a TPM value of over 6000. But the 
maximum level doesn't endure for long as it drops down very drastically. 
