---
title: "Class 9"
author: "Aisha Mohamed (A16297530)"
format: pdf
---

#Importing candy data from the Halloween Project. 


```{r}
candy <- read.csv("candy-data.txt", row.names = 1)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
fruits_candy <- as.logical(candy$fruity)
sum(fruits_candy)

```

Finding out which candies have the highest winpercent. 


> Q3. What is your favorite candy in the dataset and what is it’s 
winpercent value?

```{r}
candy["Reese's Peanut Butter cup", ]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent

```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent

```


Using `skimr` package to give a quick overview. 


```{r}
library("skimr")
skim(candy)
```


> Q6. Is there any variable/column that looks to be on a different scale to 
the majority of the other columns in the dataset?

```{r}
skim(candy$winpercent)
```

> Q7. What do you think a zero and one represent for the candy$chocolate column?

True/False.


Visually interpreting the candy dataset.  
> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)

ggplot(candy) + aes(winpercent) + geom_histogram(binwidth = 10)
```

> Q9. Is the distribution of winpercent values symmetrical?
 
 No the distribution of winpercent values is not symmetrical. 


> Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```


> Q11. On average is chocolate candy higher or lower ranked than fruit candy?


```{r}
choc.inds <- as.logical(candy$chocolate)
choc.win <- candy[choc.inds, "winpercent"]
choc.win
```

```{r}
fruit.inds <- as.logical(candy$fruity)
fruit.win <- candy[fruit.inds, "winpercent"]
fruit.win
```

```{r}
#Average of chocolate candy ranks
mean(choc.win)

#Average of fruity candy ranks 
mean(fruit.win)
```


> Q12. Is this difference statistically significant?


```{r}
t.test(choc.win, fruit.win)
```


#Overall candy rankings. 


> Q13. What are the five least liked candy types in this set?

```{r}
library(dplyr)

candy %>% 
  arrange(winpercent) %>% 
  head(5)
```


> Q14. What are the top 5 all time favorite candy types out of this set?


```{r}
candy %>% 
  arrange(desc(winpercent)) %>% 
  head(5)
```


> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) + aes(winpercent, rownames(candy)) + geom_col()
```


> Q16. This is quite ugly, use the reorder() function to get the bars sorted
by winpercent?

```{r}
ggplot(candy) + aes(winpercent, reorder(rownames(candy), winpercent)) + geom_col()
```


#Time to add some useful color. 

```{r}
my_cols = rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "red"
my_cols[as.logical(candy$fruity)] = "pink"
```


```{r}
ggplot(candy) + aes(winpercent, reorder(rownames(candy), winpercent)) + 
                      geom_col(fill = my_cols)

```


> Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is sixlets. 

> Q18. What is the best ranked fruity candy?

The best rank fruity candy is starburst.


```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 15)

```


#Taking a look at pricepercent. 


> Q19. Which candy type is the highest ranked in terms of winpercent for the 
least money - i.e. offers the most bang for your buck?


```{r}
ords <- order(candy$winpercent, decreasing = TRUE)
head( candy[ords,c(11,12)], n=5 )
```
Reece's miniatures is highest ranked in winpercent for the least amount of money. 



> Q20. What are the top 5 most expensive candy types in the dataset and of 
these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```
Nik L Nip is the the most expensive and least liked candy in the data. 



#Exploring Correlation Structure


```{r}
library(corrplot)


cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated 
(i.e. have minus values)?

The most anti-correlated two variables are 'fruity' and 'chocolate'.

> Q23. Similarly, what two variables are most positively correlated?

The most positively correlated two variables are 'chocolate' and 'bar'. 


#Principal Analysis Component


```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1], pca$x[,2], xlab = "PC1", ylab = 
       "PC2")
```

Change plotting character and add color. 

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)

```


```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])


p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```


```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 15)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other 
       (light brown), fruity (red), other (black)",
       caption="Data from 538")
```


```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> Q24. What original variables are picked up strongly by PC1 in the positive 
direction? Do these make sense to you?

Fruity, hard, and pluribus are the variables that have the most positive 
PC1 contribution. This does make sense as these variables all have high correlation 
with each other. 