---
title: "Class5"
author: "A16297530"
format: pdf
---

## Using ggplot2

The ggplot2 package needs to be installed as it does not come with R 'out of the box'

We use the `install.packages()` function to do this. 

```{r}
head(cars)
```

To use ggplot I need to load it up before I can call any of the functions in the package. I do this with the `library()` function. 

```{r}
library(ggplot2)
ggplot()
```

All ggplot figures have at least 3 things: 
- data (the stuff we want to plot)
- aesthetic mapping (aes values)
- geoms

```{r}
ggplot(cars) + aes(x=speed, y=dist) + geom_point() + labs(title = "Speed and Stopping Distances of Cars", x = "Speed (MPH)", y = "Stopping Distance (ft)", subtitle = "Your informative subtitile text here", caption = "Dataset: 'cars'") + geom_smooth(method= "lm", se = FALSE) + theme_bw()
```

ggplot is not the only graphing system in R there are lots of others. There is even "base R" graphics. 

```{r}
ggplot(cars)
```



```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

```{r}
p <- ggplot(genes) + aes(x = Condition1, y = Condition2, col = State) + geom_point() 

p 

p + scale_colour_manual(values = c("blue", "gray", "red")) + labs(title = "Gene Expression Changes Upon Drug Treatment", x = "Control (no drug)", y = "Drug Treatment")
```



```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
library(gapminder)
library(dplyr)

gapminder_2007 <- gapminder %>% filter(year==2007)

ggplot(gapminder_2007) + aes(x = gdpPercap, y = lifeExp, color = continent, size = pop) + geom_point(alpha = 0.5) +
  scale_size_area(max_size = 10) + facet_wrap(2007)


gapminder_1957 <- gapminder %>% filter(year == 1957)

ggplot(gapminder_1957) + aes(x = gdpPercap, y = lifeExp, color = continent, size = pop) + geom_point(alpha = 0.7) + scale_size_area(max_size = 10) + facet_wrap(1957)


```



```{r}
library(gapminder)
library(dplyr)


gapminder_top5 <- gapminder %>% filter(year ==2007) %>% arrange(desc(pop)) %>% top_n(5, pop)

gapminder_top5

ggplot(gapminder_top5) + geom_col(aes(x= country, y = pop))

ggplot(gapminder_top5) + aes(x = reorder(country, -pop), y = pop, fill = country) + geom_col(col = "gray30") + guides(fill = "none")

```

```{r}
library(gapminder)
library(dplyr)


head(USArrests)

USArrests$State <- rownames(USArrests)


ggplot(USArrests) + aes(x = reorder(State, Murder), y = Murder) + geom_point() + geom_segment(aes(x = State, xend = State, y = 0, yend = Murder), color = "blue") + coord_flip()
```


